Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-07-03 workflow/workflow.Rmd
Changes to scone description + zinbwave() function

```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
library(BiocStyle)
library(knitr)
knitr::opts_chunk$set(cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5)

# Bioconductor:
library(clusterExperiment)
library(BiocParallel)
library(zinbwave)
# GitHub
library(slingshot)
#library(scone)
# CRAN
library(doParallel)
library(RColorBrewer)
library(gam)
set.seed(20)
if(packageVersion("zinbwave")<'0.99.4.2') stop("must have current develop version to avoid bugs")
if(packageVersion("clusterExperiment")<'1.3.1.9003') stop("must have current develop version to avoid bugs")
```

```{r parallel, echo=FALSE}
runZinb <- TRUE
runClus <- TRUE
NCORES <- 2
mysystem = Sys.info()[['sysname']]
if (mysystem == 'Darwin'){
  registerDoParallel(NCORES)
  register(DoparParam())
}else if (mysystem == 'Linux'){
  register(bpstart(MulticoreParam(workers=NCORES)))
}else{
  print('Please change this to allow parallel computing')
  register(SerialParam())
}
```

@@

- solve loading packages issues (DR),
- versions of packages we want to use in develop branches and change functions in text accordingly (EP and DR),
- update clustering section with new clusterExperiment functions (EP),
- add details in section preprocessing (DR),
- address Sandrine's comments in sections slingshot and de (KS),
- clean up singlecellworkflow package (FP),
- update NAMESPACE and DESCRIPTION (FP),
- install and run package using cache instead of saved files for zinbwave and clusterExperiment and making sure we can build the vignette using R 3.4 and bioc-devel 3.6 (all?),
- make singlecellworkflow package public (FP).]

@@

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages; and (4) DE analysis along lineages. Throughout the workflow, we use a unique `SummarizedExperiment` object to create an easy to use sequence of steps, where a different R package is used for each step.

```{r schema, echo=FALSE, out.width='90%', fig.cap="Workflow for analyzing scRNA-seq datasets. On the right, main plots generated by the workflow."}
knitr::include_graphics('schema_workflow.png')
```

@@

```{r stemcelldiff, echo=FALSE, out.width='60%', fig.align='center', fig.cap = 'Stem cell differentiation in the mouse olfactory epithelium. This figure was reproduced with kind permission from Fletcher et al. (2017).'}
knitr::include_graphics('stemcelldiff_Fletcher2017_2e.png')
```

@@

@@

We remove the ERCC spike-in sequences and the CreER gene, as the later corresponds to a marker gene for HBCs. Davide, is that correct?

@@

Using the Bioconductor R package `scone`, we remove low-quality cells that do not pass our quality control filter, that is ... See Figure \@ref(fig:scone). Davide, would you like to describe your choices for scone here?

```{r scone, fig.cap="Plot generated by scone to filter low-quality cells.", eval=FALSE}
# QC-metric-based sample-filtering
data("housekeeping")
hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]

@@

```{r sconeFilt, eval=FALSE}
# Simplify to a single logical
mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
se <- se[, mfilt]
dim(se)
```

@@

```{r,eval=FALSE}
# Filtering to top 1000 most variable genes
vars <- rowVars(log1p(assay(se)))
names(vars) <- rownames(se)
vars <- sort(vars, decreasing = TRUE)
core <- se[names(vars)[1:1000],]
```

```{r}
load("../data/oe_se_1000Var.rda")
colnames(colData(core))[colnames(colData(core)) == 'clusterLabels'] = 'publishedClusters' 
```
[FP: delete this chunk once we can load scone]

@@

```{r zinbschema, echo=FALSE, out.width='95%', fig.cap="Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
knitr::include_graphics('zinb_schema.pdf')
```

@@

```{r zinb}
fn <- '../data/zinb_batch.rda'
if (runZinb & !file.exists(fn)){
  print(system.time(se <- zinbDimRed(core, K = 50, X = '~ Batch',
                                       residuals = TRUE,
                                       normalizedValues = TRUE)))
  save(se, file = fn)
}else{
  load(fn)
}
```

DR: the chunk above and the similar one for clusterExperiment are fine for now, but in the published workflow, we should just rely on the markdown cache system (the code above doesn't check for changes to the file or code -- just that a version of the file exists). FP: Yes.

@@

The function `zinbDimRed` returns normalized expression measures as deviance residuals from the fit of the ZINB-WaVE model, with user-supplied gene- and cell-level covariates. Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots). Note that in this case no low-dimensional matrix `W` is computed, as some of the biological signal of interest could be removed when computing residuals with respect to this matrix.

@@

```{r boxplotNorm, fig.cap="Boxplots of normalized expression measures (deviance residuals), color-coded by batch."}
norm_order <- norm[, order(as.numeric(batch))]
col_order <- col_batch[batch[order(as.numeric(batch))]]
boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
        border = col_order, xaxt = 'n', ylab='Expression measure')
abline(h=0)
```

@@

```{r pcanorm, fig.cap="PCA of normalized expression measures, where each point represents a cell. In the left panel, cells are color-coded by batch. In the right panel, cells are color-coded by original published clustering."}
pca <- prcomp(t(norm))
par(mfrow = c(1,2))
plot(pca$x, col = col_batch[batch], pch = 20, main = '')
plot(pca$x, col = col_clus[as.character(publishedClusters)], pch = 20, main = '')
```

@@

The `zinbDimRed` function can also be used to perform dimensionality reduction, where, in this workflow, the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`. The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance. To verify that `W` indeed captures the biological signal of interest, we display the MDS results in a scatterplot with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).

```{r mdsW, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
W <- colData(se)[, grepl('^W', colnames(colData(se)))]
W <- as.matrix(W)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(publishedClusters)], main = '',
     pch = 20, xlab = 'Component 1', ylab = 'Component 2')
legend(x = 'bottomright', legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = 'Sample')
```

@@

```{r rsec_50}
fn <- '../data/RSEC_W_combineMinSize10.rda'
if (runClus & !file.exists(fn)){
  # Symbol for samples missing from original clustering
  seObj <- SummarizedExperiment(t(W), colData = colData(core))
  print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
                                  betas = 0.8, dimReduce="none",
                clusterFunction = "hierarchical01", minSizes=1,
                ncores = NCORES, isCount=FALSE,
	       dendroReduce="none",dendroNDims=NA,
                subsampleArgs = list(resamp.num=100,
                                     clusterFunction="kmeans",
                                     clusterArgs=list(nstart=10)),
                verbose=TRUE,
                combineProportion = 0.7,
                mergeMethod = "none", random.seed=424242,
                combineMinSize = 10)))
  save(ceObj, file = fn)
}else{
  load(fn)
  colnames(colData(ceObj))[colnames(colData(ceObj)) == 'clusterLabels'] = 'publishedClusters'
}
```

FP: Same comment as Davide's comment for zinbwave. We should remove the if file exists and rely on the cache.

@@

plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
            whichClusters = "primary", clusterFeaturesData = "all",
            clusterSamplesData = "dendrogramValue", breaks = 0.99,
            sampleData = c("publishedClusters", "Batch", "Experiment"),
            clusterLegend = metaColors, annLegend = FALSE, main = '')
```

@@

```{r mdsWce, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clusters."}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, 'color']
names(pal) <- palDF[, 'name']
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = '', pch = 20,
     xlab = 'Component1', ylab = 'Component2')
legend(x = 'bottomright', legend = names(pal), cex = .5,
       fill = pal, title = 'Sample')
```

@@

```{r}
our_cl <- primaryClusterNamed(ceObj)
cl = our_cl[our_cl != "-1"]
pal = pal[names(pal) != '-1']
X <- W[our_cl != "-1", ]
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
X <- mds$points

@@

```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
```

```{r tree, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred structure."}
pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])
```

@@

```{r slingshotsupervised, eval=FALSE}
lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
                         end.clus = c("c3", "c6"))
lineagees <- getCurves(lineages)
pairs(lineages, type='curves', pal[primaryClusterNamed(ceObj)])
pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])

@@

More formally, for each lineage, we use the robust local regression method loess to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime. We then can test the null hypothesis of no change over time for each gene using the `gam` package. We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).

@@

In an effort to improve scRNA-seq data analysis workflows, we are currently exploring a variety of applications and extensions of our ZINB-WaVE model. In particular, we are developing a method to impute counts for dropouts; the imputed counts could be used in subsequent steps of the workflow, including dimensionality reduction, clustering, and cell lineage inference. ZINB-WaVE can also be used for identifying genes that are differentially expressed between cells, both in terms of the negative binomial mean and the zero inflation probability, reflecting, respectively, gradual DE and on/off DE patterns. We are also developing a method to identify genes that are DE either within or between lineages inferred from Slingshot.

>>>>

```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE}
if(packageVersion("zinbwave")<"0.99.5") stop("must have current devel version to avoid bugs")
if(packageVersion("clusterExperiment")<"1.3.1.9003") stop("must have current develop version to avoid bugs")
if(packageVersion("scone")<"1.1.2") stop("must have current devel version to avoid bugs")

library(BiocStyle)
library(knitr)
knitr::opts_chunk$set(cache=TRUE, error=FALSE, message=FALSE, warning=TRUE)
knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5)
```

@@

- solve loading packages issues (DR), DONE!
- versions of packages we want to use in develop branches and change functions in text accordingly (EP and DR), [DR: scone and zinbwave now OK!]
- update clustering section with new clusterExperiment functions (EP),
- add details in section preprocessing (DR),
- address Sandrine's comments in sections slingshot and de (KS),
- clean up singlecellworkflow package (FP),
- update NAMESPACE and DESCRIPTION (FP),
- install and run package using cache instead of saved files for zinbwave and clusterExperiment and making sure we can build the vignette using R 3.4 and bioc-devel 3.6 (all?),
- make singlecellworkflow package public (FP).]

@@

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` R package; and (4) DE analysis along lineages. Throughout the workflow, we use a unique `SummarizedExperiment` object to store the scRNA-seq data along with any metadata available from the experiment.

```{r schema, echo=FALSE, out.width="90%", fig.cap="Workflow for analyzing scRNA-seq datasets. On the right, main plots generated by the workflow."}
knitr::include_graphics("schema_workflow.png")
```

@@

```{r stemcelldiff, echo=FALSE, out.width="60%", fig.align="center", fig.cap = "Stem cell differentiation in the mouse olfactory epithelium. This figure was reproduced with kind permission from Fletcher et al. (2017)."}
knitr::include_graphics("stemcelldiff_Fletcher2017_2e.png")
```

@@

## Package versions

We need the following packages.

```{r packages}
# Bioconductor:
library(clusterExperiment)
library(BiocParallel)
library(zinbwave)
library(scone)

# GitHub
library(slingshot)

# CRAN
library(doParallel)
library(RColorBrewer)
library(gam)
set.seed(20)
```

Note that in order to successfully run the workflow, we need the devel versions
of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.5)` and `clusterExperiment (>=1.3.2)` [DR: Elizabeth, please check!]. We recommend running Bioconductor 3.6 (currently the devel version). See https://www.bioconductor.org/developers/how-to/useDevel/ on how to use the Bioconductor devel.

```{r parallel, echo=FALSE}
runZinb <- TRUE
runClus <- TRUE
NCORES <- 6
mysystem = Sys.info()[["sysname"]]
if (mysystem == "Darwin"){
  registerDoParallel(NCORES)
  register(DoparParam())
}else if (mysystem == "Linux"){
  register(bpstart(MulticoreParam(workers=NCORES)))
}else{
  print("Please change this to allow parallel computing")
  register(SerialParam())
}
```

@@

We remove the ERCC spike-in sequences and the CreER gene, as the latter corresponds to the estrogen receptor fused to Cre recombinase (Cre-ER), which is used to activate HBC into differentiation following injection of tamoxifen (see [@Fletcher2017] for details).

@@

Using the Bioconductor R package `scone`, we remove low-quality cells that do not pass our quality control filter (See Figure \@ref(fig:scone)). The function `metric_sample_filter` uses several criteria to filter out low-quality samples. First, we filter out samples with low total number of reads or low alignment percentages. Then, we look at the number of housekeeping genes that are detected in each sample, and filter out those samples with a low detection rate. See the [scone vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/scone/inst/doc/sconeTutorial.html) for details on the filtering procedure.

```{r scone, fig.cap="Plot generated by scone to filter low-quality cells."}
# QC-metric-based sample-filtering
data("housekeeping")
hk = rownames(se)[toupper(rownames(se)) %in% housekeeping$V1]

@@

```{r sconeFilt}
# Simplify to a single logical
mfilt <- !apply(simplify2array(mfilt[!is.na(mfilt)]), 1, any)
se <- se[, mfilt]
dim(se)
```

After sample filtering, we are left with `r NCOL(se)` good quality cells.

@@

```{r}
# Filtering to top 1000 most variable genes
vars <- rowVars(log1p(assay(se)))
names(vars) <- rownames(se)
vars <- sort(vars, decreasing = TRUE)
core <- se[names(vars)[1:1000],]
```

@@

```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
knitr::include_graphics("zinb_schema.pdf")
```

@@

```{r zinb}
print(system.time(se <- zinbwave(core, K = 50, X = "~ Batch",
                                 residuals = TRUE,
                                 normalizedValues = TRUE)))
```

@@

The function `zinbwave` returns a `SummarizedExperiment` object that includes normalized expression measures, defined as deviance residuals from the fit of the ZINB-WaVE model, with user-supplied gene- and cell-level covariates. Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots). Note that in this case the low-dimensional matrix `W` is not included in the residuals computation, to avoid the removal of the biological signal of interest.

[DR: Fanny, did you figure out why we get infinite values? If not, we need to at least explain why we are doing what we do below.]

@@

```{r boxplotNorm, fig.cap="Boxplots of normalized expression measures (deviance residuals), color-coded by batch."}
norm_order <- norm[, order(as.numeric(batch))]
col_order <- col_batch[batch[order(as.numeric(batch))]]
boxplot(norm_order, col = col_order, staplewex = 0, outline = 0,
        border = col_order, xaxt = "n", ylab="Expression measure")
abline(h=0)
```

@@

```{r pcanorm, fig.cap="PCA of normalized expression measures, where each point represents a cell. In the left panel, cells are color-coded by batch. In the right panel, cells are color-coded by original published clustering."}
pca <- prcomp(t(norm))
par(mfrow = c(1,2))
plot(pca$x, col = col_batch[batch], pch = 20, main = "")
plot(pca$x, col = col_clus[as.character(publishedClusters)], pch = 20, main = "")
```

@@

The `zinbwave` function can also be used to perform dimensionality reduction, where, in this workflow, the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`. The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance. To verify that `W` indeed captures the biological signal of interest, we display the MDS results in a scatterplot with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).

```{r mdsW, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by original published clustering."}
W <- colData(se)[, grepl("^W", colnames(colData(se)))]
W <- as.matrix(W)
d <- dist(W)
fit <- cmdscale(d, eig = TRUE, k = 2)
plot(fit$points, col = col_clus[as.character(publishedClusters)], main = "",
     pch = 20, xlab = "Component 1", ylab = "Component 2")
legend(x = "bottomright", legend = unique(names(col_clus)), cex = .5,
       fill = unique(col_clus), title = "Sample")
```

@@

```{r rsec_50}
seObj <- SummarizedExperiment(t(W), colData = colData(core))
print(system.time(ceObj <- RSEC(seObj, k0s = 4:15, alphas = c(0.1),
                                betas = 0.8, dimReduce="none",
                                clusterFunction = "hierarchical01", minSizes=1,
                                ncores = NCORES, isCount=FALSE,
                                dendroReduce="none",dendroNDims=NA,
                                subsampleArgs = list(resamp.num=100,
                                                     clusterFunction="kmeans",
                                                     clusterArgs=list(nstart=10)),
                                verbose=TRUE,
                                combineProportion = 0.7,
                                mergeMethod = "none", random.seed=424242,
                                combineMinSize = 10)))
```

@@

plotHeatmap(ceObj, visualizeData = assays(se)$normalizedValues,
            whichClusters = "primary", clusterFeaturesData = "all",
            clusterSamplesData = "dendrogramValue", breaks = 0.99,
            sampleData = c("publishedClusters", "Batch", "Experiment"),
            clusterLegend = metaColors, annLegend = FALSE, main = "")
```

@@

```{r mdsWce, fig.cap="MDS of the low-dimensional matrix W, where each point represents a cell and cells are color-coded by RSEC clusters."}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, "color"]
names(pal) <- palDF[, "name"]
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20,
     xlab = "Component1", ylab = "Component2")
legend(x = "bottomright", legend = names(pal), cex = .5,
       fill = pal, title = "Sample")
```

@@

```{r}
our_cl <- primaryClusterNamed(ceObj)
cl = our_cl[our_cl != "-1"]
pal = pal[names(pal) != "-1"]
X <- W[our_cl != "-1", ]
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
X <- mds$points

@@

```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
pairs(lineages, type="curves", pal[primaryClusterNamed(ceObj)])
```

```{r tree, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred structure."}
pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)])
```

@@

```{r slingshotsupervised, eval=FALSE}
lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
                         end.clus = c("c3", "c6"))
lineagees <- getCurves(lineages)
pairs(lineages, type="curves", pal[primaryClusterNamed(ceObj)])
pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)])

@@

More formally, for each lineage, we use the robust local regression method `loess` to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime. We then can test the null hypothesis of no change over time for each gene using the `gam` package. We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes ordered by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).

@@

In an effort to improve scRNA-seq data analysis workflows, we are currently exploring a variety of applications and extensions of our ZINB-WaVE model. In particular, we are developing a method to impute counts for dropouts; the imputed counts could be used in subsequent steps of the workflow, including dimensionality reduction, clustering, and cell lineage inference. In addition, we are extending ZINB-WaVE to differentially expression, both in terms of the negative binomial mean and the zero inflation probability, reflecting, respectively, gradual DE and on/off DE patterns. We are also developing a method to identify genes that are DE either within or between lineages inferred from Slingshot.
